Nature Index product

### 2023-10-31 DRAFT_insectIndicators.Rmd
added GDM calculation, GDE remains

_Author and date:_
Jens Åström `r Sys.Date()`

@@

This workflow is shortened by putting most of the underlying code in separate R-packages, which are freely available through Github. The relevant packages are `bombLepiSurv` https://github.com/jenast/bombLepiSurv for the bumblebee and butterfly indicators (**Not public yet**), and `Norimon` https://github.com/jenast/Norimon for the rest of the indicators.

@@

knitr::kable(indList)
```
Many of these indicators are self-explanatory and don't require much justification. **Bumblebees in semi-natural grasslands** and **Butterflies in semi-natural grasslands** are functionally important pollinators, and are as a community dependent on a varied herbaceous flora. They therefore both represent key functionality, as well as indicators of habitat quality. **Biomass of flying insects** represent a key functional characteristic in ecosystem functioning as food for insectivores, and as a response to habitat quality and management. **Species richness of flying insects** represents the total breadth of the insect community, and **Species richness of pollinating insects** complements the butterfly and bumblebee indicators in taxonomic breadth and spatial resolution.
**Species richness of dung associated insects** represent a key functional group in decomposition, and function as an indicator of grazing intensity and diversity. **Quotent of Apocrita vs Symphyta species richness** tries to go beyond simple amounts, and capture compositional changes in the insect community. Apocrita covers predatory and parasitic wasps, while Symphyta is plant-feeding wasps. A relative decrease in predatory/parasitic wasps could indicate a degradation in insect prey communities, and a decrease in the potential for natural pest control. **Number of alien species** simply sums the number of alien species found in a region. This is based on the alien species list from the Norwegian Biodiversity Information Center, which lists known alien species and an increasing list of door-knocker species. Note that due to large historical knowledge gaps of insect species distributions, there could potentially be many more alien species in Norway, that are not picked up on this alien species list. Lastly, **Intraspecific genetic variation** captures biodiversity that until now has been hidden, due to lack of practical technologies to measure it. The need to conserve genetic diversity is identified in Norwegian law, and decreased intraspecific diversity could signal population decreases, and decreased potential to adapt to environmental changes, before whole species are lost to an area.

@@

## References
* Åström, Jens; Birkemoe, Tone; Brandsegg, Hege; Dahle, Sondre; Davey, Marie Louise; Ekrem, Torbjørn; Fossøy, Frode; Hanssen, Oddvar; Laugsand, Arne Endre; Majaneva, Markus; Staverløkk, Arnstein; Sverdrup-Thygeson, Anne; Ødegaard, Frode.
Insektovervåking på Østlandet, Sørlandet og i Trøndelag. Rapport fra feltsesong 2022. Trondheim: Norsk institutt for naturforskning (NINA) 2023 (ISBN 978‐82‐426‐5037‐5) 99 s. NINA rapport(2241)

@@

* Connor M French, Laura D Bertola, Ana C Carnaval, Evan P Economo, Jamie M Kass, David J Lohman, Katharine A Marske, Rudolf Meier, Isaac Overcast, Andrew J. Rominger, Phillip Staniczenko, Michael J Hickerson. 2023. Global determinants of insect mitochondrial genetic diversity. https://doi.org/10.1101/2022.02.09.479762

@@

Note that the output currently contains some experimental values, awaiting a more robust methodology. "Shannon diversity" doesn't really make sense without counts or amounts, etc. This will be replaced by measurements of genetic diversity that are based on published peer-reviewed methodology (Connor et al. 2023), in the (hopefully near) future.

@@

@@

@@

```{r}
plot(biomass_sn_diff)
```

@@

biomass_sn_to_exp <- biomass_sn_to_exp %>% 
  select(year_window,
         region = region_name,
         ind_value = boot_value,
         ind_sd = boot_sd,
         ind_lower2.5 = boot_lower2.5,
         ind_upper97.5 = boot_upper97.5)

@@

```{r}
plot(richness_sn_diff)
```

@@

richness_sn_to_exp <- richness_sn_to_exp %>% 
  select(year_window,
         region = region_name,
         ind_value = boot_value,
         ind_sd = boot_sd,
         ind_lower2.5 = boot_lower2.5,
         ind_upper97.5 = boot_upper97.5)

@@

```{r}
plot(pollinators_sn_diff)
```

@@

pollinators_sn_to_exp <- pollinators_sn_to_exp %>% 
  select(year_window,
         region = region_name,
         ind_value = boot_value,
         ind_sd = boot_sd,
         ind_lower2.5 = boot_lower2.5,
         ind_upper97.5 = boot_upper97.5)

@@

```{r}
plot(dung_sn_diff)
```

@@

dung_sn_to_exp <- dung_sn_to_exp %>% 
  select(year_window,
         region = region_name,
         ind_value = boot_value,
         ind_sd = boot_sd,
         ind_lower2.5 = boot_lower2.5,
         ind_upper97.5 = boot_upper97.5)

@@

par_sym_sn_exp <- par_sym_sn_exp %>% 
  select(year_window,
         region = region_name,
         ind_value = boot_value,
         ind_sd = boot_sd,
         ind_lower2.5 = boot_lower2.5,
         ind_upper97.5 = boot_upper97.5)

@@

#### Number of alien species in semi-natural land
Here we measure the number of species found that are present on the alien species list from the Norwegian biodiversity information center.

Fetch the species list.
```{r}
alien_list <- tbl(con,
                  Id(schema = "lookup",
                     table = "fremmedartslista_2018_artsdatabanken"))

```
```{r}
alien_list
```

```{r}
alien_species <- alien_list %>% 
  select(species = scientificName) %>% 
  pull()
```
```{r}
alien_sn <- get_observations(subset_year = 2020:2022,
                               subset_habitat = "Semi-nat",
                               subset_species = alien_species)
alien_sn
```

Bootstrap the alien observations
```{r}
alien_sn_boot <- bootstrap_value(alien_sn,
                                   value = no_species,
                                   groups = c("year",
                                              "region_name"))

@@

Compare to reference value and normalize. [Here we need 0 as reference value. How to normalize? No upper bound.] Use the proportion of species? That gives very low values

<!-- ```{r} -->
<!-- alien_spec_ref <- 0 -->

<!-- alien_sn_diff <- alien_sn_boot  -->
<!-- ``` -->

```{r}
plot(alien_spec_boot)
```

```{r}
alien_sn_exp <- alien_sn_diff$bootstrap_summary

alien_sn_exp <- alien_sn_exp %>% 
  select(year_window,
         region = region_name,
         ind_value = boot_value,
         ind_sd = boot_sd,
         ind_lower2.5 = boot_lower2.5,
         ind_upper97.5 = boot_upper97.5)

alien_sn_exp <- norway_reg_NorIns  %>% 
  inner_join(alien_sn_exp,
            by = c("region" = "region"),
            multiple = "all")

alien_sn_exp
```

#### Genetic diversity evenness of flying insects in semi-natural land
This is not fully implemented yet, and the method shown here functions as a placeholder, awaiting a proper calculation. Here, the intraspecific variation is based simply on the number of unique genetic variants (ASVs), and their relative abundance measured as number of DNA-reads in the samples. But to do this properly, we need to use the full COI sequence to calculate the true genetic diversity (GD), and from there the mean diversity (GDM) and the evenness (GDE) of a set of species in a spatial and temporal unit  (see Connor et al. 2023). This work is on the way.

@@

```{r}
plot(gde_sn_diff)
```

@@

gde_sn_to_exp <- gde_sn_to_exp %>% 
  select(year_window,
         region = region_name,
         ind_value = boot_value,
         ind_sd = boot_sd,
         ind_lower2.5 = boot_lower2.5,
         ind_upper97.5 = boot_upper97.5)

>>>>

_Author and date:_
Jens Åström, Marie Davey `r Sys.Date()`

@@

This workflow is shortened by putting most of the underlying code in separate R-packages, which are freely available through Github. The relevant packages are `bombLepiSurv` https://github.com/jenast/bombLepiSurv for the bumblebee and butterfly indicators, and `Norimon` https://github.com/jenast/Norimon for the rest of the indicators.

@@

knitr::kable(indList)
```
Many of these indicators are self-explanatory and don't require much justification. **Bumblebees in semi-natural grasslands** and **Butterflies in semi-natural grasslands** are functionally important pollinators, and are as a community dependent on a varied herbaceous flora. They therefore both represent key functionality, as well as indicators of habitat quality. **Biomass of flying insects** represent a key functional characteristic in ecosystem functioning as food for insectivores, and as a response to habitat quality and management. **Species richness of flying insects** represents the total breadth of the insect community, and **Species richness of pollinating insects** complements the butterfly and bumblebee indicators in taxonomic breadth and spatial resolution.
**Species richness of dung associated insects** represent a key functional group in decomposition, and function as an indicator of grazing intensity and diversity. **Quotent of Apocrita vs Symphyta species richness** tries to go beyond simple amounts, and capture compositional changes in the insect community. Apocrita covers predatory and parasitic wasps, while Symphyta is plant-feeding wasps. A relative decrease in predatory/parasitic wasps could indicate a degradation in insect prey communities, and a decrease in the potential for natural pest control. **Number of alien species** takes into account the number of alien species found in a region and optionally, the frequency of sampling events they are observed in. This is based on the alien species list from the Norwegian Biodiversity Information Center, which lists known alien species and an increasing list of door-knocker species. Note that due to large historical knowledge gaps of insect species distributions, there could potentially be many more alien species in Norway, that are not picked up on this alien species list. Lastly, **Intraspecific genetic variation** captures biodiversity that until now has been hidden, due to lack of practical technologies to measure it. The need to conserve genetic diversity is identified in Norwegian law, and decreased intraspecific diversity could signal population decreases, and decreased potential to adapt to environmental changes, before whole species are lost to an area.

@@

## References

* French, C. M., Bertola, L. D., Carnaval, A. C., Economo, E. P., Kass, J. M., Lohman, D. J., … Hickerson, M. J. 2023, juli 17. Global determinants of insect mitochondrial genetic diversity. bioRxiv. bioRxiv. https://doi.org/10.1101/2022.02.09.479762

* Miraldo, A., Li, S., Borregaard, M. K., Flórez-Rodríguez, A., Gopalakrishnan, S., Rizvanovic, M., … Nogués-Bravo, D. 2016. An Anthropocene map of genetic diversity. Science 353(6307): 1532–1535. https://doi.org/10.1126/science.aaf4381

* Rohde, K. 1992. Latitudinal Gradients in Species Diversity: The Search for the Primary Cause. Oikos 65(3): 514. https://doi.org/10.2307/3545569

* Theodoridis, S., Fordham, D. A., Brown, S. C., Li, S., Rahbek, C., & Nogues-Bravo, D. 2020. Evolutionary history and past climate change shape the distribution of genetic diversity in terrestrial mammals. Nature Communications 11(1): 2557. https://doi.org/10.1038/s41467-020-16449-5

* Åström, Jens; Birkemoe, Tone; Brandsegg, Hege; Dahle, Sondre; Davey, Marie Louise; Ekrem, Torbjørn; Fossøy, Frode; Hanssen, Oddvar; Laugsand, Arne Endre; Majaneva, Markus; Staverløkk, Arnstein; Sverdrup-Thygeson, Anne; Ødegaard, Frode.
Insektovervåking på Østlandet, Sørlandet og i Trøndelag. Rapport fra feltsesong 2022. Trondheim: Norsk institutt for naturforskning (NINA) 2023 (ISBN 978‐82‐426‐5037‐5) 99 s. NINA rapport(2241)

@@

@@

Note that the output currently contains some experimental values, awaiting a more robust methodology. "Shannon diversity" doesn't really make sense without counts or amounts, etc. This will be replaced by measurements of genetic diversity that are based on published peer-reviewed methodology (French et al. 2023), in the (hopefully near) future.

@@

#### Time-series plots
The default plots are good for showing the uncertainty distributions, but can be a bit difficult to track the time series trends with. We can use the `ts_plot` function for a simple time-series plot.

```{r}
ts_plot(poll_richness_boot)
```

@@

The averaging over 5-years naturally also shows up in the time-series plots. Note the very wide confidence intervals of the Sørlandet region, with only 1-year worth of data.

```{r}
ts_plot(poll_richness_boot_5_year)
```

@@

```{r biomass_diff}
plot(biomass_sn_diff)
```

@@

biomass_sn_to_exp <- biomass_sn_to_exp %>% 
  select(year,
         region = region_name,
         ind_value = boot_value,
         ind_sd = boot_sd,
         ind_lower2.5 = boot_lower2.5,
         ind_upper97.5 = boot_upper97.5)

@@

```{r}
plot(richness_sn_diff,
     limit = c(0, 1))
```

```{r richness}
map_plot(richness_sn_diff,
         whole_country = T,
         limit = c(0, 1))
```

@@

richness_sn_to_exp <- richness_sn_to_exp %>% 
  select(year,
         region = region_name,
         ind_value = boot_value,
         ind_sd = boot_sd,
         ind_lower2.5 = boot_lower2.5,
         ind_upper97.5 = boot_upper97.5)

@@

```{r}
plot(pollinators_sn_diff,
     limit = c(0, 1))
```

@@

pollinators_sn_to_exp <- pollinators_sn_to_exp %>% 
  select(year,
         region = region_name,
         ind_value = boot_value,
         ind_sd = boot_sd,
         ind_lower2.5 = boot_lower2.5,
         ind_upper97.5 = boot_upper97.5)

@@

```{r}
plot(dung_sn_diff,
     limit = c(0, 1))
```

@@

dung_sn_to_exp <- dung_sn_to_exp %>% 
  select(year,
         region = region_name,
         ind_value = boot_value,
         ind_sd = boot_sd,
         ind_lower2.5 = boot_lower2.5,
         ind_upper97.5 = boot_upper97.5)

@@

par_sym_sn_exp <- par_sym_sn_exp %>% 
  select(year,
         region = region_name,
         ind_value = boot_value,
         ind_sd = boot_sd,
         ind_lower2.5 = boot_lower2.5,
         ind_upper97.5 = boot_upper97.5)

@@

#### Number of alien species in semi-natural land
Here we measure the prevalence of alien species in the trap catches. We present two options for calculating the indicator, as it is not clear at this point which of them is superior.

The simplest approach is to just sum up the number of alien species present in a region, i.e. the species richness of alien species. This can be related to the total species richness of the community, and expressed as the proportion of alien species in the community. The reference state of alien species is zero by definition, since we don't want any alien species. The value for good ecological condition could be set at a low (but arbitrary) proportion of alien species in the community. This is done for alien plant species where the value for good ecological condition is set to 0.05 (< 5 % of the plant community should be alien). For a hyper-diverse community such as insects, this is probably too high a value. In a community of say 20 000 insect species, 5 % would mean 1 000 alien species, which is hard to see as a good condition. It is not clear to us at present what level to set here, but we exemplify the calculations with a good condition set to 0.1 % (< 20 species in a 20 000 species community). It is also likely that the species richness of the entire community will increase, perhaps substantially, with increased sampling. It is not clear if the number of detected native and alien species will increase at the same rate, so it may be that the proportion of alien species will change, perhaps substantially, with increased sampling effort. It may be difficult therefore to interpret this frequency of alien insect species.

In addition, this approach will not take into account the spread of the species, a measure that is especially relevant for rare but increasing species. A species found only once will count as much as a species found in every locality. To take into account the prevalence of the species, we can observe the proportion of localities each alien species is observed in. These proportions could then be summed together for at total score of alien species. Here, a species found in all localities will count much more, than a species only found once.

For each alien species, the observation frequency is calculated as the number of localities the species is observed in, divided by the total amount of localities in a geographical region and time period. $$p_i = \frac{1}{N} \sum_{i=1, j=1}^{N}{obs_{ij}},$$ where $obs_ij$ is an observation of species $i$ in locality $j$, and $N$ is the total number of possible observations (localities in the region and time period).
These proportions are combined into a common score for all alien species simply by summing them together; $$alien~score =\sum_{i=1}^{S}{~p_{i}},$$ where $p_i$ is the observation frequency of species $i$, and $S$ is the set of all observed alien species. The uncertainty of both types of indicators is calculated by bootstrapping the set of localities within a temporal and spatial scope, similarly to the other indicators. If all alien species are present in all localities, the indicator value simply becomes $S$. If the species is not present in all localities, it's influence on the indicator is weighted down linearly with it presence frequency.

**Fetch the species list.**

The alien species observations are stored in a materialized view for convenience. This is simply a list of observations of the alien species of categories LO, PH, HI, or SE, together with the year localities and locality samplings they were observed in. For the purposes of this calculation, we will only consider the year localities (and not the prevalence og alien species within locality and season). So the first thing to do is to fetch this list, and aggregate it to the year locality level. This summary is based on the 2023 of the alien species evaluation, and we only consider species that are identified with identification confidence "Moderate" or "High" (not Low or Poor). In addition, we filter out the forest sites.

```{r}
alien_obs <- tbl(con,
                 Id(schema = "views",
                    table = "alien_obs")) %>% 
  filter(identification_confidence %in% c("MODERATE", "HIGH"),
         grepl("Semi-nat", locality))
```

```{r}
alien_obs_ls <- alien_obs %>% 
  summarize(as.integer(n_distinct(ls_id))) %>% 
  pull()

alien_obs_yl <- alien_obs %>% 
  summarize(as.integer(n_distinct(yl_id))) %>% 
  pull()

no_alien_spec <- alien_obs %>% 
  summarize(as.integer(n_distinct(species_latin_fixed))) %>% 
  pull()

no_alien_cat <- alien_obs %>% 
  summarize(as.integer(n_distinct(risikokategori_2023))) %>% 
  pull()

@@

We can have quick look at some statistics of the data. We see that the project thus far has identified `r no_alien_spec`, representing `r no_alien_cat` risk-categories, in (all) `r alien_obs_yl` semi-natural localities.

To continue, we aggregate the presences of alien species to locality years, meaning in recording in how many localities each alien species were detected at least once.

```{r}
alien_obs_yl <- alien_obs %>% 
  group_by(yl_id,
           locality,
           year,
           region_name,
           risikokategori_2023,
           species_latin_fixed) %>% 
  summarize(present = sum(present) > 0) %>% 
  arrange(locality, species_latin_fixed) %>% 
  collect(na.rm = TRUE)
```

```{r}
alien_obs_yl
```

Next step is to bootstrap the alien observations to get the uncertainty measurement. This has to be done in a slightly different way than for the aggregated diversity or biomass values that we have used for the other indicators. Here we have a set of alien species presences for a locality, and when we bootstrap the localities, all these presences should go together. So we first have to take bootstrap samples of sets of localities (with replacement). Then we calculate the presence (or presence frequency) of each alien species $p_i$, before we can summarize all of these into the indicator value $H_{alien}$. This is implemented in a separate function `bootstrap_alien_obs` in the `Norimon` package, were we can optionally either sum the species numbers, or sum the species presence frequencies. Note that for all alien species recorded in at least one locality, we have recorded the presences and absences in all localities. This is needed for his function to work. This function is more computer intensive than the bootstrap of already aggregated values, so we employ parallelization through the `parallel` package internally.

**We first calculate the indicator based on the the number of alien species**

```{r}
alien_sn_boot <- bootstrap_alien_obs(alien_obs_yl,
                                     groups = c("year",
                                                "region_name"),
                                     rolling_year_window = TRUE,
                                     only_no_spec = TRUE)

```{r}
plot(alien_sn_boot)
```

The jagged lines in the density plot comes from the discrete values of number of species. Østlandet has the most alien species, followed by Sørlandet, and lastly Trøndelag. Trøndelag has a fewer total number of species, but similar amount of alien species in each locality, leading to a very narrow confidence band from the bootstrap.

```{r}
ts_plot(alien_sn_boot)
```

As explained above, these values could be expressed as a proportion of the the total species richness. Since the total species richness is different for the different regions, we (currently) have to do this separately for each region.

```{r}
alien_obs_yl_sor <- alien_obs_yl %>% 
  filter(region_name == "Sørlandet")

alien_obs_yl_ost <- alien_obs_yl %>% 
  filter(region_name == "Østlandet")

alien_obs_yl_trond <- alien_obs_yl %>% 
  filter(region_name == "Trøndelag")

```{r}
alien_sn_sor_boot <- bootstrap_alien_obs(alien_obs_yl_sor,
                                     groups = c("year",
                                                "region_name"),
                                     rolling_year_window = TRUE,
                                     only_no_spec = TRUE)
alien_sn_ost_boot <- bootstrap_alien_obs(alien_obs_yl_ost,
                                     groups = c("year",
                                                "region_name"),
                                     rolling_year_window = TRUE,
                                     only_no_spec = TRUE)
alien_sn_trond_boot <- bootstrap_alien_obs(alien_obs_yl_trond,
                                     groups = c("year",
                                                "region_name"),
                                     rolling_year_window = TRUE,
                                     only_no_spec = TRUE)
```

Next, we fetch a record of the species richness found so far, in each region.

```{r}
region_richn <- get_observations(agg_level = "region_habitat")
```
```{r}
region_richn
```

```{r}
region_richn_sor <- region_richn %>% 
  filter(region_name == "Sørlandet",
         habitat_type == "Semi-nat") %>% 
  select(no_species) %>% 
  pull() %>% 
  as.numeric()

region_richn_ost <- region_richn %>% 
  filter(region_name == "Østlandet",
         habitat_type == "Semi-nat") %>% 
  select(no_species) %>% 
  pull() %>% 
  as.numeric()

region_richn_trond <- region_richn %>% 
  filter(region_name == "Trøndelag",
         habitat_type == "Semi-nat") %>% 
  select(no_species) %>% 
  pull() %>% 
  as.numeric()

```
Next, we scale each indicator to the total number of observed species.

```{r}
alien_sn_sor_prop <- alien_sn_sor_boot / region_richn_sor
alien_sn_ost_prop <- alien_sn_ost_boot / region_richn_ost
alien_sn_trond_prop <- alien_sn_trond_boot / region_richn_trond
```

We could scale these values with a reference value of 0, and a breaking point of 0.001. eaTools has a function for that, but for this to work, we currently can only operate on the aggregate values of the indicator (not the bootstrap samples).

```{r}
alien_sn_prop <- alien_sn_sor_prop[[1]] %>% 
  rbind(alien_sn_ost_prop[[1]]) %>% 
  rbind(alien_sn_trond_prop[[1]]) %>% 
  select(year,
         region = region_name,
         alien_score = boot_value)
```
```{r}
alien_sn_prop_sf <- norway_reg_NorIns  %>% 
  inner_join(alien_sn_prop,
            by = c("region" = "region"),
            multiple = "all")
```

```{r}
alien_sn_prop_sf$alien_score_scaled <- eaTools::ea_normalise(alien_sn_prop_sf, 
                                              vector = "alien_score", 
                                              upper_reference_level = 0.01,
                                              lower_reference_level = 0,
                                              break_point = 0.001, 
                                              reverse = TRUE,
                                              plot = FALSE)
```

This calculation is then an the sf-object can be exported.

**Secondly, we calculate the indicator based on the distribution of the alien species**

We call this alien_H, since the calculation resembles Hill numbers from calculations of Shannon information.

```{r}
alien_H_boot <- bootstrap_alien_obs(alien_obs_yl,
                                    groups = c("year",
                                               "region_name"),
                                    rolling_year_window = TRUE,
                                    only_no_spec = FALSE)

```
```{r}
alien_H_boot
```

```{r}
plot(alien_H_boot)
```
Interstingly, by this calculation method, the different regions have much more similar values. Specifically, the southern region has comparable values to the eastern region, while Trøndelag has lower values. This indicates that region Østlandet has markedly higher number of alien species than the Southern region, but with generally more limited distribution. While in comparison, the alien species in the south are more frequently distributed throughout the sampling localities.

```{r}
ts_plot(alien_H_boot)
```

This indicator also has a natural reference value, with zero presences of alien species, meaning an alien score of 0. We could scale this indicator similarly as with the indicator based on the pure number of alien species. Similar to that case, there is no upper limit to how many alien species there can be, but we set a high enough value, and use a breaking point for the scaling, this upper end point has limited influence.

We don't have enough experience to determine an appropriate breakpoint for the scaling at this point, or an upper value, but exemplify the calculation with some subjective values

alien_H_boot_sf <- alien_H_boot[[1]] %>% 
  select(year,
         region = region_name,
         alien_score = boot_value)

alien_H_boot_sf <- norway_reg_NorIns  %>% 
  inner_join(alien_H_boot_sf,
            by = c("region" = "region"),
            multiple = "all")

alien_H_boot_sf$alien_score_scaled <- eaTools::ea_normalise(alien_H_boot_sf,
                                                            vector = "alien_score", 
                                                            upper_reference_level = 10,
                                                            lower_reference_level = 0,
                                                            break_point = 2, 
                                                            reverse = TRUE,
                                                            plot = FALSE)

This is then ready for export.
```{r}
alien_H_boot_sf
```

#### Genetic diversity of flying insects in semi-natural land
The genetic diversity (GD) of a species is a measurement of the differences between DNA variants within the same species, here calculated on the COI region of the mitochondrial DNA. It is in principle a different measurement than the total number of species in a community, or the number of distinct DNA variants in a community. It is an empirical question whether these measurement correlate correlate on the global or local scale. For instance, Miraldo et al. (2016) found global patterns of decreasing GD near the poles, coinciding with global patterns of species richness. These patterns may share causes, such as higher mutation rates and shorter generation times in warmer climates (Rohde K. 1992). However, it is not clear if and how these patterns are distributed within a smaller region, such as within Norway. Further, spatial and temporal trends in GD, may can indicate pressures on species, before they are visible as loss of species.

Following e.g. Miranda et al. (2016), we define GD as the average number of nucleotide differences per nucleotide site across all pairwise sequence comparisons within a species. Mathematically this can be expressed as: $$\hat{\prod}_s = \frac{1}{(\frac{n}{2})} \sum_{i=1}^{n-1}\sum_{j=i+1}^{n}\frac{k_{ij}}{m_{ij}},$$ where $s$ is a single species, $k_{ij}$ is the number of different nucleotides between sequence $i$ and sequence $j$, $(\frac{n}{2})$ represents the number of pairwise comparisons made, and $m_{ij}$ is the number of shared base pairs between sequence $i$ and $j$ (Theodoridis et al 2020). Following French et al. (2023), we here considered only species that where found at least 3 separate times, to make sure we had enough samples to observe any intraspecific diversity.

This species-specific measurement can be combined into a community measurement in several ways. Firstly, we can simply take the average GD of a community of species (GDM), or $$GDM = \frac{1}{S}\sum_{s=1}^{S}\hat{\prod}_s, $$ where $S$ is the total number of species, and $s$ is each species (Theodoridis et al. 2020). The mean genetic diversity of a community is a natural starting point, but can - analogous to species richness -, hide important patterns in the relative contributions from individual species. In short, the distributions of GD might be skewed, so that a few species carries the most influence on the measurement. Therefore, French et al. (2023) has developed a genetic evenness measurement (GDE), that is analogous to Shannon's entropy for communities on the species level. This is defined as $$\frac{exp(\sum_{i=0}^{N} - \hat{\prod}_i ln(\hat{\prod}_i))}{N},$$ where N is the number of species in the community, and $\hat{\prod}_i$ is the GD calculated above for a single species. The numerator is the Hill number of the Shannon entropy. This results in a measurement that can be compared across communities of different species richness.

We here present both GDM and GDE, as potential indicators. These are relatively new measures, and it is the first time it is calculated for large insect communities in Norway. We need further experience with these measurements to tease out potential pro's and con's in the Norwegian setting. The raw-data on the DNA-sequences is as of yet not  publicly available, and the calculations of GDM and GDE is not shown in detail.

**NOTE** As of now, we only have calculations by Marie on GDM, for all species on the year locality level. The GDE still remains to be calculated. We here import the ready made calculations for file.

```{r}
gdm <- readRDS("data/GDM_year_locality.rds")
```
The `gdm` contains the mean genetic diversity for the species found in every year locality combination. This is calculated after first filtering on xxx, xxxx.

The `gdm` object contains the year_locality_ids as a key, which we can join to the relevant metadata, e.g. through the alien observation object used above.

```{r}
gdm_yl <- alien_obs_yl  %>% 
  ungroup() %>% 
  select(yl_id, locality, year, region_name) %>% 
  distinct() %>% 
  inner_join(gdm, by = c("yl_id"= "year_locality_id"))

gdm_yl
```

Since these calculations are so novel, it can be worth spending some time exploring how the mean genetic diversity relates to other richness measurement at the locality level. Remember that globally, the species richness and genetic mean diversity show a general positive correlation (Miraldo et al. 2016).

```{r}
richness_yl <- get_observations(agg_level = "year_locality",
                                subset_habitat = "Semi-nat")
```

```{r}
richness_gdm_join <- richness_yl %>% 
  inner_join(gdm_yl,
             by = c("locality" = "locality",
                    "year" = "year",
                    "region_name" = "region_name"))
```

```{r}
ggplot(richness_gdm_join) +
  geom_point(aes(x = no_species,
                 y = GDM,
                 color = region_name)) +
  geom_smooth(aes(x = no_species,
                 y = GDM,
                 color = region_name),
              method = lm) +
  scale_color_nina(name = "Region") +
  xlab("Antall arter")
```
Interestingly, and perhaps reassuringly, the number of species doesn't seem to be clearly related to the average genetic diversity. We can look at the mean number of ASVs as well.

```{r}
ggplot(richness_gdm_join) +
  geom_point(aes(x = mean_no_asv_per_species,
                 y = GDM,
                 color = region_name)) +
  geom_smooth(aes(x = mean_no_asv_per_species,
                 y = GDM,
                 color = region_name),
              method = lm) +
  scale_color_nina(name = "Region") +
  xlab("Antall genetiske varianter per art")
```
Here we see a positive relationship in Trøndelag (although uncertain), but not the other regions.

```{r}
ggplot(richness_gdm_join) +
  geom_point(aes(x = no_species,
                 y = mean_no_asv_per_species,
                 color = region_name)) +
  geom_smooth(aes(x = no_species,
                 y = mean_no_asv_per_species,
                 color = region_name),
              method = lm) +
  scale_color_nina(name = "Region") +
  xlab("Antall arter") +
  ylab("Antall genetiske varianter per art")

```
While the number of ASVs (genetic variants) and the number of species per locality show more clear relationships. This was just a quick look at the genetic aspect of biodiversity, that require some further analysis to support robust interpretations. In the meantime, we continue with the calculation example of the indicators. We bootstrap the mean genetic diversity like with the other indicators.

```{r}
gdm_yl_boot <- bootstrap_value(df = gdm_yl,
                                   value = GDM,
                                   groups = c("year",
                                              "region_name")
                                   )
gdm_yl_boot
```

```{r}
plot(gdm_yl_boot)
```

So far, Trøndelag seem to have lower genetic diversity than Sørlandet and Østlandet.

```{r}
ts_plot(gdm_yl_boot)
```

Although, as with the other indicators, we see variation between years. This can be seen if we do not bootstrap on a rolling 5 year window.

```{r}
gdm_yl_yearly_boot <- bootstrap_value(df = gdm_yl,
                                      value = GDM,
                                      groups = c("year",
                                                 "region_name"),
                                      rolling_year_window = FALSE
                                   )
```

```{r}
ts_plot(gdm_yl_yearly_boot)

Below is some placeholder calculation, in wait of the proper GDE calculations.

@@

```{r}
plot(gde_sn_diff,
     limit = c(0, 1))
```

@@

gde_sn_to_exp <- gde_sn_to_exp %>% 
  select(year,
         region = region_name,
         ind_value = boot_value,
         ind_sd = boot_sd,
         ind_lower2.5 = boot_lower2.5,
         ind_upper97.5 = boot_upper97.5)
